---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# type

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/EthanSansom/type/graph/badge.svg)](https://app.codecov.io/gh/EthanSansom/type)
<!-- badges: end -->

{type} provides a lightweight type system for R, meant for users to quickly and
easily add run-time type validation to functions and variables. The {type} 
package supports:

- Adding type annotations to function arguments and return values.
- Declaring typed variables and constants.
- Defining new types and type aliases.

```{r, include = FALSE}
library(type)
library(purrr)
```

## Installation

You can install the development version of {type} from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("EthanSansom/type")
```

## Types

A "type" in {type} defines a set of restrictions used to validate other objects 
in R. For example, the type `t_bool` (for "boolean type") requires that an object:

- Is a bare logical vector.
- Is scalar (size 1).
- Contains no `NA` values.

In other words, an object of type `t_bool` must be either `TRUE` or `FALSE`. 

When a type is printed, it displays the requirements that an object of that 
type must meet:

```{r}
print(t_bool)
```

`obj_is_type()` may be used to test whether an object meets the requirements
of a given type.

```{r}
obj_is_type(10L, t_bool)
obj_is_type(TRUE, t_bool)
```

For additional information on why a type test failed or succeeded, use 
`obj_inspect_type()` to print a diagnostic message.

```{r}
obj_inspect_type(NA, t_bool)
```

A variable may be restricted to a given type using the `%:%` operator:

```{r}
# Require that `x` is an integer, and initialize `x` to `1:5`
t_int %:% x(1:5)
x 
```

Once a variable is typed, it may be re-assigned to a variable of the same type,
but cannot be assigned a different type:

```{r}
x <- 2L
x

x <- "A"
```

Alternatively, `obj_assert_type()` may be used to interactively enforce the 
type of an object:

```{r}
obj_assert_type(data.frame(x = 1:3), t_int)
```

Both `%:%`, `obj_assert_type()`, and other type validation functions in {type} 
raise {rlang} errors of class `<type_error_mistyped>` when an object is mistyped.

### Base Types

{type} includes the following base types for common R objects:

| Object      | Matches                                          |
|-------------|--------------------------------------------------|
| `t_any`     | Any object                                       |
| `t_null`    | `NULL`                                           |
| `t_list`    | A list                                           |
| `t_env`     | An environment                                   |
| `t_fun`     | A function                                       |
| `t_vec`     | A {vctrs} style vector                           |
| `t_num`     | An numeric (e.g. integer or double) vector       |
| `t_lgl`     | A bare logical vector                            |
| `t_bool`    | A single `TRUE` or `FALSE`                       |
| `t_int`     | A bare integer vector                            |
| `t_dbl`     | A bare double vector                             |
| `t_chr`     | A bare character vector                          |
| `t_string`  | A single non-`NA` string                         |
| `t_dataframe` | A data frame                                   |
| `t_factor`  | A factor                                         |
| `t_date`    | A `Date`                                         |
| `t_posixct` | A `POSIXct` datetime                             |

### Traits

Every type is built using traits, which add restrictions to a type. For example,
the base `t_string` type is built using the `sized()` and `complete()` traits:

```{r}
# A string is:
t_string <- t_chr |> # A character vector
  sized(1L) |>       # Of size 1
  complete()         # Which is non-missing
```

Traits compose naturally using the `|>` operator, making it easy to define
custom types.

```{r}
t_probability <- t_dbl |> bounded(0, 1, "[]")

obj_inspect_type(0.5, t_probability)
```

### Structural Types

{type} defines a small dialect for setting type constraints on parts of an
object using `on()` selector functions, `has()`, and `has_relation()`:

```{r}
# `names()` of an object must contain "x" and "y"
t_any |> has(on(names), t_chr |> contains(c("x", "y")))

# Element `[[1]]` of an object must be a function
t_any |> has(on_elm(1L), t_fun)

# Every element of the object must be the same size
t_list |> has_relation(same_sized(on_each()))
```

These are useful for building complex types from a simple set of traits.

```{r}
t_coordinates <- t_list |>
  has(on(names), t_chr |> same_as(c("lat", "lon"))) |>
  has(on_elm("lat"), t_dbl |> bounded(-90, 90)) |>
  has(on_elm("lon"), t_dbl |> bounded(-180, 180)) |>
  has_relation(same_sized(on_elm("lat"), on_elm("lon")))
 
good <- list(lat = c(51.5, 40.7), lon = c(-0.1, -74.0))
bad <- list(lat = c(51.5, 200.0), lon = c(-0.1, -74.0, 0.0))
 
obj_is_type(good, t_coordinates)
obj_inspect_type(bad, t_coordinates)
```

`list_type()` and `dataframe_type()` provide a shorthand for typed containers:

```{r}
t_survey <- dataframe_type(
  id = t_int |> complete(), 
  score = t_dbl |> bounded(0, 100)
)
print(t_survey)
```

While `list_of_type()` constructs the common case of a homogenous list type:

```{r}
t_list_of_chr <- list_of_type(t_chr)
obj_is_type(list("A", "B", "C"), t_list_of_chr)
obj_is_type(list("A", 2L, "C"), t_list_of_chr)
```

### Constants

A variable may be declared a constant using the `const()` modifier, in which
case it's value cannot be changed after assignment.

```{r}
const(t_int) %:% z(1L)
z <- 2L
```

## Typed Functions

